Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes and Reservoirs

### 2019-08-08 manuscript/manuscript.Rmd
Very close to internal review draft!

As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable.  Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings [e.g. @NEED_CITATION].  However, in many lakes there are still likely sufficient nutrients present, as a result of eutrophication, to allow for increases in productivity.  Additionally, these results point to the fact that productivity is driven by processes operating at different scales.  For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].  Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as catchment percent impervious and lake morphology can also impact temperature [@nelson2007stream]. In short, differences in regional and state level trends are driven by complex and multi-scale processes.

@@

There are several possible approaches to further reduce nutrient loads. First, nutrient load reductions may be possible through source controls and enhanced entrainment and treatment of ground and surface waters transporting nutrients to receiving waters.  Green infrastructure approaches are one way to possibly achieve both goals [e.g. @NEED_CITATION].  Additionally, within-lake approaches may also be used.  A promising approach is the restoration of freshwater mussels to waterbodies that historically had those species. Some studies have shown reductions in both nutrients and algal biomass [@kreeger2018restoration].

@@

The analysis approach we used here, site-specific z-scores, is not a novel method, but using it to examine water quality trends is a new application of the technique, as we could find few examples of using it specifically for water quality trends [@NEED_CITATION].  This approach, and in particular non-scaled anomalies, does have a long history in the analysis of trends in climate [@hansen2006global; hansen2010global; @jones1996calculating; @jones1999surface].  We built on these methods and adapted them for use with long-term water quality trends.  While other methods are valid and robust [e.g. @Oliver_2017], we chose averaged site-specific z-scores as they can provide readily interpretable results, especially for communicating to general audiences.  In addition, the site-specific z-scores are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (e.g., added later in a time period or removed).  Lastly, this analysis is only possible because of the availability of sound, long-term data on water quality in Rhode Island.  Without the URIWW data and the commitment and participation of more than 2500 volunteers over the years, our analyses would have been impossible.  Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in capturing and better understanding long term environmental trends.

>>>>

As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable.  Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent.  However, in many lakes there are still likely sufficient nutrients present, as a result of eutrophication, to allow for increases in productivity.  Additionally, these results point to the fact that productivity is driven by processes operating at different scales.  For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].  Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as catchment percent impervious and lake morphology can also impact temperature [@nelson2007stream]. In short, differences in regional and state level trends are driven by complex and multi-scale processes.

@@

There are several possible approaches to further reduce nutrient loads [@yang2018nutrients]. First, nutrient load reductions may be possible through source controls and enhanced entrainment and treatment of ground and surface waters transporting nutrients to receiving waters.  Green infrastructure approaches are one way to possibly achieve both goals [@pennino2016watershed; @reisinger2019changes].  Additionally, within-lake approaches may also be used.  Another approach is the restoration of freshwater mussels to waterbodies that historically had those species. Some studies have shown reductions in both nutrients and algal biomass [@kreeger2018restoration].

@@

The analysis approach we used here, site-specific z-scores, is not a novel method, but using it to examine water quality trends is a new application of the technique, as we could find little evidence of using it specifically for water quality trends.  This approach, and in particular non-scaled anomalies, does have a long history in the analysis of trends in climate [@hansen2006global; hansen2010global; @jones1996calculating; @jones1999surface].  We built on these methods and adapted them for use with long-term water quality trends.  While other methods are valid and robust [e.g. @Oliver_2017], we chose averaged site-specific z-scores as they can provide readily interpretable results, especially for communicating to general audiences.  In addition, the site-specific z-scores are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (e.g., added later in a time period or removed).  Lastly, this analysis is only possible because of the availability of sound, long-term data on water quality in Rhode Island.  Without the URIWW data and the commitment and participation of more than 2500 volunteers over the years, our analyses would have been impossible.  Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in capturing and better understanding long term environmental trends.
